\documentclass{letter}
%\usepackage[top=2cm, bottom=2cm, inner=2cm, outer=2cm]{geometry}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=teal,
    filecolor=teal,
    citecolor=black,      
    urlcolor=teal,
}
    
\title{utrecht}
\author{Emilio Berti}
\date{August 2023}

\begin{document}

\textbf{References}

\underline{Prof. Jens-Christian Svenning}\\
Section for Ecoinformatics and Biodiversity\\
Aarhus University\\
Aarhus, Denmark.

email: svenning@bio.au.dk

phone: +45 289 92 304

Prof. Svenning was my PhD supervisor and has always supported my personal and professional growth.
We have worked together on several publications and still maintain active collaborations.

\vspace{1cm}

\underline{Prof. Ulrich Brose}\\
Theory in Biodiversity Science\\
German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig\\
Leipzig, Saxony, Germany.

email: ulrich.brose@idiv.de

phone: +49 341 9733205

Prof. Brose is my current PostDoc supervisor and he has actively helped me to further develop my career goals and objsectives.

\vspace{1cm}

\underline{Caitlin Wilkinson}\\
Theory in Biodiversity Science\\
German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig\\
Leipzig, Saxony, Germany.

email: caitlin.wilkinson@idiv.de

phone: +49 341 9739178

Wilkinson is a PhD student who I have been tutoring since approximately one year.
She accepted to be a reference for this application and you are encouraged to contact her to inquire about my approach and commitment to teaching and mentoring.

\end{document}
