---
title: "What are ENM/SDM and why do we use them?"
author: "Emilio Berti"
format:
  beamer:
    engine: tectonic
    colortheme: beaver
    incremental: false
    aspectratio: 43
---

## The why SDMs

Species Distribution Models (SDMs) are projections of species geographic ranges, i.e. where the species are found.

Among other applications, SDMs are used for:

  1. Understand what limits species' distribution.
  2. Project how they will change under future climate.
  3. Plan conservation and management, e.g. where to establish new protected areas.

## A brief history of SDMs

  - 1807: Essay on the Geography of Plants (Humboldt).
  - 1967: The theory of island biogeography (MacArthur & Wilson)
  - 1990s: Infancy of quantitative SDMs.
  - 2000s: SDMs as a recognized field.
  - 2010s: Boom of SDMs.
  - 2020s: Overflow of SDMs.

## The basic workflow

Starting from data, to obtain the *projected* distribution of one species:

  1. Ecological Niche Model (ENM): Model the ecological requirements of species.
  2. SDM: Project these requirements into geographic space.

Always Data $\rightarrow$ ENM $\rightarrow$ SDM.

## The basic workflow

Always Data $\rightarrow$ ENM $\rightarrow$ SDM.

Given some data, ENM is a flexible module. 

SDM is usually much more constrained: Meta-population dynamics and other spatial processes.

## The basic workflow

Always Data $\rightarrow$ ENM $\rightarrow$ SDM.

ENM: Environmental processes.
ENM is hard due to data limitations.

SDM: Spatial processes
SDM is hard due to complexity and lack of theory.

# Theory Primer

## Ecological Niche Modeling (ENM)

ENM try to reconstruct the ecological (climatic) niche of a species.

Theory suggests the niche must be convex.

```{r niche-symm, echo=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
l <- function(x, mu, sigl, sigr) {
  out <- rep(NA, length(x))
  out[x < mu] <- exp(- ((x[x < mu] - mu) / sqrt(2) / sigl) ^ 2)
  out[x >= mu] <- exp(- ((x[x >= mu] - mu) / sqrt(2) / sigr) ^ 2)
  return(out)
}
x <- seq(-10, 40, length.out = 1e3)
par(mfrow = c(1, 1), mar = c(4, 4, 2, 2))
plot(
  x, l(x, 10, 5, 5),
  type = "l", lwd = 2, col = "dodgerblue3",
  xlab = "Temperature", 
  ylab = "Suitability"
)
```

## Ecological Niche Modeling (ENM)

Theory suggests the niche must be convex and not symmetrical

```{r niche-asymm, echo=FALSE, fig.width=4, fig.height=3, fig.align='center'}
x <- seq(-10, 40, length.out = 1e3)
par(mfrow = c(1, 1), mar = c(4, 4, 2, 2))
plot(
  x, l(x, 10, 5, 10),
  type = "l", lwd = 2, col = "dodgerblue3",
  xlab = "Temperature", 
  ylab = "Suitability"
)
```

## Ecological Niche Modeling (ENM)

Example of non-symmetrical, convex 2D niche.

```{r niche-2d, echo=FALSE, fig.width=4, fig.height=4, fig.align='center'}
x <- seq(-10, 40, length.out = 1e2)
y <- seq(300, 5000, length.out = 1e2)
z <- matrix(NA, nrow = length(unique(x)), ncol = length(unique(y)))
for (X in unique(x)) {
  for (Y in unique(y)) {
    i <- which(unique(x) == X)
    j <- which(unique(y) == Y)
    z[i, j] <- l(X, 10, 5, 12) * l(Y, 2000, 500, 1e3)
  }
}
image(z, axes = FALSE, col = hcl.colors(100, "Fall", rev = TRUE))
contour(z, add = TRUE, levels = seq(0.1, 1, by = 0.2))
axis(1, at = seq(0, 1, length.out = 6), labels = seq(-10, 40, length.out = 6))
axis(2, at = seq(0, 1, length.out = 6), labels = seq(300, 5000, length.out = 6))
title(xlab = "Temperature", ylab = "Precipitation")
```

## Species Distribution Modeling (SDM)

SDMs project ENMs from an environmental space to a geographical area.

```{r sdm, echo=FALSE, fig.width=4, fig.height=3, fig.align='center'}
w <- terra::vect("../data/ne_110m_land.shp")
w <- terra::crop(w, terra::ext(-13, 33, 33, 62))
p <- terra::vect(
  matrix(c(
    4.3226, 49.1915,
    1.6996, 47.9762,
    0.7767, 47.2537,
    1.2625, 45.3487,
    3.3025, 44.7575,
    5.6341, 44.7903,
    6.4598, 46.1698,
    9.3742, 47.4179,
    2.7744, 47.1551,
    5.8345, 47.3522,
    7.8260, 50.2754,
    4.1769, 49.9798,
    4.3226, 49.1915
  ), ncol = 2, byrow = TRUE),
  type = "points", crs = terra::crs(w)
)
p <- terra::convHull(p)
p <- terra::densify(p, interval = 1e3)
par(mfrow = c(1, 2), mar = c(4, 4, 2, 2))
layout(matrix(c(1, 1, 2), ncol = 3))
image(z, axes = FALSE, col = hcl.colors(100, "Fall", rev = TRUE))
contour(z, add = TRUE, levels = seq(0.1, 1, by = 0.2))
axis(1, at = seq(0, 1, length.out = 6), labels = seq(-10, 40, length.out = 6))
axis(2, at = seq(0, 1, length.out = 6), labels = seq(300, 5000, length.out = 6))
title(xlab = "Temperature", ylab = "Precipitation")
terra::plot(
  w, fun = \() terra::polys(p, col = "green3"),
  mar = c(0, 0, 0, 0)
)
```
