---
title: "Data for ENMs/SDMs"
subtitle: ""
author: ""
format:
  beamer:
    engine: tectonic
    colortheme: beaver
    incremental: false
    aspectratio: 169
---

## Two types of data

  - Species' occurrence data
  - Environmental (climate) data


# Species' occurrence data

## Species' occurrence data

  - Where the species was observed (`1`) and where not (`0`)
  - Typically, only where the species was observed

  1. GBIF: \url{https://www.gbif.org/}
  2. iNaturalist: \url{https://www.inaturalist.org/}
  3. eBird: \url{https://ebird.org/}
  4. \dots

## Species' occurrence data

Strong geographic and taxonomic biases

::::{.columns}

:::{.column width=60%}
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
library(terra)
r <- rast("../data/effort-mammalia.tif")
r[r > 2e9] <- NA
l <- vect("../data/coastline.shp")
plot(
  log10(r + 1),
  col = hcl.colors(100, "Spectral", rev = TRUE),
  maxcell = 1e5,
  fun = \() lines(l),
  frame = FALSE, axes = FALSE
)
```
\centering
Number of mammal records ($log_{10}$)
:::

:::{.column width=40%}
 - Mostly animals
 - Mostly vertebrates
 - Mostly birds and mammals
:::

::::

## GBIF

 - Global Biodiversity Information Facility (GBIF)
 - $\approx 3.5$ billion records
 - Direct upload and pull from other databases

Go to \url{https://www.gbif.org/} to download data

# Climate data

## Climate data

 1. WorldClim: https://www.worldclim.org/
 2. CHELSA: https://chelsa-climate.org/
 3. PRISM: https://www.prism.oregonstate.edu/
 4. \dots

\pause

Every climate data is different and you should use the one that is best for your case.

In this course, we will use WorldClim, which is easy to use, very reliable, and often used in real applications.

Go to \url{https://www.worldclim.org/} to download bioclimatic variables at 10 arc-minute resolution.

## Bioclimatic variables

  - Generated from temperature and precipitation
  - Annual values, averaged over $\approx$ 30 years
  - Usually, 19 variables, but sometimes more

\pause

For example:

  1. Mean annual temperature (BIO01)
  2. Maximum temperature of the warmest month (BIO05)
  3. Total annual precipitation (BIO12)
  4. Precipitation seasonality (BIO15)

## BIO01

```{r, echo=FALSE}
r <- rast("../data/wc2.1_10m_bio_1.tif")
plot(r, col = hcl.colors(100, "Spectral", rev = TRUE), frame = FALSE, axes = FALSE)
```
