---
title: "Running the first ENM/SDM"
author: ""
format:
  beamer:
    engine: tectonic
    colortheme: beaver
    incremental: false
    aspectratio: 169
---

# ENM

## Data

```{r load data}
library(mgcv)
library(gratia)

d <- read.csv("../data/occurrences.csv")
head(d, n = 3)
```

## Correlation among variables

```{r, correlation, echo=FALSE, fig.height=3.2, fig.width=3.2, fig.align='center'}
image(
  1:8, 1:8,
  cor(d[, grepl("wc2", colnames(d))]),
  zlim = c(-1, 1),
  col = c(hcl.colors(99, "RdBu", rev = TRUE), "grey20"),
  xlab = "",
  ylab = "",
  useRaster = TRUE, axes = FALSE
)
grid(nx = 8, ny = 8, lty = 1, col = "grey20", lw = 2)
axis(
  1,
  at = 1:8,
  labels = gsub("wc2.1_10m_bio_", "BIO", colnames(d)[grepl("wc2", colnames(d))]),
  lwd = 2,
  las = 2
)
axis(
  2,
  at = 1:8,
  labels = gsub("wc2.1_10m_bio_", "BIO", colnames(d)[grepl("wc2", colnames(d))]),
  lwd = 2,
  las = 2
)
for (i in 1:4) {
  axis(
    i,
    at = c(0, 9),
    labels = NA,
    lwd = 2
  )
}
```

 - Bioclimatic variables are highly correlated with each others.
 - A subset of them should be used for ENM/SDM.

## Selecting variables

Variable selection

 - The usual statistical tricks ...
 - Informed by the **biology** of the species.

Because I do not know the biology of the species, I will use a statistical trick: Akaike information criterion (AIC).

## Full ENM

```{r full-enm}
enm <- gam(
  occ ~ s(wc2.1_10m_bio_1, k = 5) +
    s(wc2.1_10m_bio_4, k = 5) + 
    s(wc2.1_10m_bio_12, k = 20) +
    s(wc2.1_10m_bio_15, k = 5),
  data = d,
  family = "binomial"
)
k.check(enm)
```

## Inspect ENM

```{r inspect-full-enm, fig.width=3.5, fig.height=3.5, fig.align='center', echo=FALSE}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 2))
for (s in unlist(smooth_terms(enm))) {
  plot(
    smooth_estimates(enm)[[s]],
    plogis(smooth_estimates(enm)[[".estimate"]]),
    type = "l", main = s, xlab = "climate", ylab = "P(X = 1)"
  )  
}
par(mfrow = c(1, 1))
```

# SDM

## Prepare rasters

Load layers.

```{r sdm-load}
library(terra)

ff <- paste0("../data/", unlist(smooth_terms(enm)), ".tif")
r <- rast(ff)
roi <- ext(-13, 33, 33, 62) # roi of Europe
r <- crop(r, roi) # crop to Europe
```

## Project distribution

```{r sdm-predict, fig.width=3.5, fig.height=3.5, fig.align='center'}
sdm <- predict(r, enm, type = "response")
writeRaster(sdm, "../data/hist-sdm.tif", overwrite = TRUE)
plot(sdm, col = hcl.colors(100, "Spectral", rev = TRUE))
```

## Binary projection

If we are interested in a binary map, e.g. showing the climatic range of the species, we need to binarize this continuous value into `0/1`.

True skill statistics (TSS).

 1. Pick a threshold value `th`.
 2. Set all cells $<th$ to `0`.
 3. Set all cells $\geq th$ to `w`.
 4. Calculate the TSS.
 5. Repeat for a different value of `th`.
 6. Report maximum TSS and use `th` for which $TSS = \max(TSS)$.

$TSS = \frac{TP\,\times\,TN - FP\,\times\,FN}{(TP+FN)(TN+FP)}$

## Binary projection

```{r sdm-binary, fig.width=4, fig.height=4, fig.align='center'}
suit <- extract(sdm, d[, c("x", "y")], ID = FALSE)[, 1]
threshold <- seq(0.1, 0.9, by = 0.001)
tss <- rep(NA, length(threshold)) # empty vector for storage
for (i in seq_along(threshold)) {
  p <- ifelse(suit > threshold[i], 1, 0)
  TP <- sum(p == 1 & d$occ == 1)
  FP <- sum(p == 1 & d$occ == 0)
  FN <- sum(p == 0 & d$occ == 1)
  TN <- sum(p == 0 & d$occ == 0)
  tss[i] <- TP / (TP + FN) + TN / (TN + FP) - 1
}
th <- threshold[which.max(tss)]  # best threshold
sdm_bin <- ifel(sdm >= th, 1, 0) # binarize map
```

## Binary projection

```{r sdm-binary-plot, fig.width=4, fig.height=4, fig.align='center', echo=FALSE}
plot(
  sdm, col = hcl.colors(100, "Spectral", rev = TRUE),
  fun = \() {
    plot(
      sdm_bin,
      col = c("#22222290", "#22222200"),
      add = TRUE, legend = FALSE
    )
  }
)
```

## Binary projection

Our map is quite *incorrect* for this species: <https://www.iucnredlist.org/species/61550/12514105>

Why do we get such bad projections compared to the known range from IUCN?

We will answer this in a next lecture.

## Hint: geographic biases in the detections

```{r sdm-issue, fig.width=3.5, fig.height=3.5, fig.align='center', echo=FALSE}
plot(
  sdm_bin,
  col = c("grey90", "firebrick"),
  colNA = "skyblue",
  legend = FALSE,
  fun = \() {
    points(d[d$occ == 1, c("x", "y")], col = "grey20", pch = 3)
  }
)
```
