---
title: "Future projections"
author: ""
format:
  beamer:
    engine: tectonic
    colortheme: beaver
    incremental: false
    aspectratio: 169
---

## Future climate data

[WorldClim data](https://www.worldclim.org/data/cmip6/cmip6_clim10m.html)

\pause

 0. Periods: 2021-2040, 2041-2060, 2061-2080, 2081-2100
 1. **G**lobal **C**irculation **M**odel (GCM) = climate model
 2. **S**hared **S**ocio-economic **P**athway (SSP) = geopolitical model

\ pause

SSP3: *A resurgent nationalism, concerns about competitiveness and security, and regional conflicts push countries to increasingly focus on domestic or, at most, regional issues.*

## Load data

```{r load}
library(terra)
library(mgcv)

enm <- readRDS("../data/enm.rds")
enm$formula

r <- rast("../data/wc2.1_10m_bioc_MPI-ESM1-2-HR_ssp370_2081-2100.tif")
r <- r[[c("bio01", "bio04", "bio12", "bio15")]]
names(r) <- paste0("wc2.1_10m_bio_", c(1, 4, 12, 15))
```

## Predict

```{r predict, fig.width=3.5, fig.height=3.5, fig.align='center'}
sdm_future <- predict(r, enm, type = "response")
plot(sdm_future, col = hcl.colors(100, "Spectral", rev = TRUE))
```

## Compare

```{r compare, fig.width=3.5, fig.height=3.5, fig.align='center'}
sdm <- rast("../data/hist-sdm.tif")
comp <- c(sdm, sdm_future)
names(comp) <- c("Historical", "MPI:SSP3:2081-2100")
panel(comp, col = hcl.colors(100, "Spectral", rev = TRUE))
```

## Change

```{r change, fig.width=3.5, fig.height=3.5, fig.align='center', echo=FALSE}
delta <- sdm_future - sdm
plot(
  delta,
  col = colorRampPalette(c("darkred", "white", "seagreen"))(100),
  range = c(-.91, .91),
  colNA = "skyblue",
  fun = \() {
    l <- delta
    l[!is.na(l)] <- 1
    lines(as.polygons(l))
  }
)
```

## Projections

 - Predictions guess what will happen in the future
 - Projections are model-based extrapolations
 \pause 
 - Climate is hard to model $\rightarrow$ Many GCMs
 \pause
 - The future is uncertain $\rightarrow$ Many SSPs

\pause

You should run projections for many combinations of GCMs and SSPs and ensemble the results.
