---
title: "Species Distribution Models"
subtitle: "An introductory course at SFU Jena"
author: "Emilio Berti"
---

## Course overview

This course introduces Ecological Niche Modeling (ENM) and Species Distribution Modeling (SDM) using the R programming language.

::: {.callout-important}
This course assumes basic R knowledge.
:::

In the first \# days, etc.

::: {.callout-note}
A basic knowledge of Geographic Information Systems is required for ENM/SDM.
I will introduce basic concepts and operations of of GIS, but this is not a course about GIS.

I lecture a GIS course every year at iDiv in Leipzig: [GIS course](https://emilio-berti.github.io/teaching/gis-ydiv/main.html).
SFU Jena also offers GIS courses: [MSc Environmental and Georesources Management ](https://www.chemgeo.uni-jena.de/en/56067/teaching), [MSc Geoinformatic](https://www.uni-jena.de/6655/m-sc-geoinformatik), and [BSc Geography](https://www.uni-jena.de/7195/b-sc-geographie).
:::

## Learning objectives

- Understand the theory behind ENM/SDM.
- Perform ENM/SDM in R.
- Apply ENM/SDM to real data.
- Produce maps.

## Software

We will use two extremely popular R package: `terra` and `mgcv`.
For `terra`, you might want to install also `codetools`, which performs checks on the C++ code of `terra`.

```{r, eval=FALSE}
install.packages(c("codetools", "terra"))
```

For `mgcv`, you might want to install also `gratia`, which provides functionality to plot and inspect the output of `mgcv`.

```{r, eval=FALSE}
install.packages(c("mgcv", "gratia"))
```

::: {.callout-note}
`terra` is still in active development and some functionalities change with newer releases.
:::
