---
title: "Running our first ENM/SDM"
author: "Emilio Berti"
format:
  html: 
    code-fold: show
  pdf: default
---

We can now model the ecological niche and the distribution of the species _Podarcis muralis_.

## Species niche

We start by loading the species and climate data that we prepared in the previous section.

```{r load data}
d <- read.csv("../data/occurrences.csv")
head(d)
```

The column names `wc2.1_10m_bio_<id>` means that this is data from WorldClim (`wc`) at a resolution of 10 arc-minutes (`10m`) for bioclimatic variables (`<id>`).
Bioclimatic variables are generally highly correlated with each others and only a subset of them should be used for train an ecological niche model.
Variable selection can be performed with the usual statistical tricks or, even better, can be informed by the biology of the species.
Because I do not know the biology of the species, I build several competing ENMs and test which one is best using the Akaike information criterion (AIC).

The first model considers only average temperature (`BIO01`) and total precipitation (`BIO12`) to train an ENM using `glm()`.

```{r first-glm}
enm_01_12 <- glm(
  occ ~ poly(wc2.1_10m_bio_1, 2, raw = TRUE) + poly(wc2.1_10m_bio_12, 2, raw = TRUE),
  data = d,
  family = "binomial"
)
```

We can test how well these variables explain the distribution of the species by building another ENM with different variables and comparing it with the model above.
For example, we can use the minimum temperature of the coldest month (`BIO06`) and the precipitation of the driest month (`BIO14`) instead.

```{r second-glm}
enm_06_14 <- glm(
  occ ~ poly(wc2.1_10m_bio_6, 2, raw = TRUE) + poly(wc2.1_10m_bio_14, 2, raw = TRUE),
  data = d,
  family = "binomial"
)
```

We can compared the two models by AIC, with the best most having the lowest AIC.

```{r aic}
AIC(enm_01_12, enm_06_14)
```

The model with the second set of variables explain the distribution of the species better than the first model.

### Statistical dredging (optional and dangerous)

There are packages to run each possible model starting from a full model and rank them according to AIC.
The most popular package is `MuMIn`, which contains the function `dredge()` to perform this steps.
Start with the full model, i.e. the most complex model, and pass this to `dredge()`.
```{r dredge}
# full model
full_enm <- glm(
  occ ~ poly(wc2.1_10m_bio_1, 2, raw = TRUE) + 
   poly(wc2.1_10m_bio_12, 2, raw = TRUE) + 
   poly(wc2.1_10m_bio_13, 2, raw = TRUE) + 
   poly(wc2.1_10m_bio_14, 2, raw = TRUE) + 
   poly(wc2.1_10m_bio_15, 2, raw = TRUE) + 
   poly(wc2.1_10m_bio_5, 2, raw = TRUE) + 
   poly(wc2.1_10m_bio_6, 2, raw = TRUE),
  data = d,
  family = "binomial"
)

# dredging
options(na.action = "na.fail")   # required by MuMIn, don't think about it
dr <- MuMIn::dredge(full_enm)
head(dr)
```

This table contains all possible sub-models of `full_enm` (including also `full_enm`) as rows ranked by AICc.
A model contains the variable if there is a `+` at that column.
The column `weight` quantifies the support of that model.
We see that the model in first row (lowest AICc) is strongly supported against all others and we can select only that as our ENM.
This model is actually the full model as it contains all bioclimatic variables specified.

## Species distribution

When using `glm()`, `terra` makes it extremely easy to produce a SDM.
We first need to load the raster layers of the bioclimatic variables.

```{r sdm-load}
library(terra)

ff <- list.files("../data", pattern = ".tif") # all files with .tif extension
r <- rast(file.path("..", "data", ff))
roi <- ext(-13, 33, 33, 62) # roi of Europe
r <- crop(r, roi) # crop to Europe
```

The `terra` function `predict(<raster>, model)` is all we need.

```{r sdm-predict-dummy, fig.width=5, fig.height=4, fig.align='center', eval=FALSE}
sdm <- predict(r, full_enm, type = "response")
plot(sdm, col = hcl.colors(100, "Spectral", rev = TRUE))
```

```{r sdm-predict, fig.width=5, fig.height=4, fig.align='center', echo=FALSE}
ff <- list.files("../data", pattern = ".tif") # all files with .tif extension
r <- rast(file.path("..", "data", ff))
roi <- ext(-13, 33, 33, 62) # roi of Europe
r <- crop(r, roi) # crop to Europe
sdm <- predict(r, full_enm, type = "response")
plot(sdm, col = hcl.colors(100, "Spectral", rev = TRUE))
```

### Binary projections

And we obtained the projected suitability of the species for Europe.
Note that this is a continuous value, in this case representing the probability of detecting the species given climate.
If we are interested in a binary map, e.g. showing the climatic range of the species, we need to binarize this continuous value into `0/1`.
There are several approaches to achieve this, but here we consider only the approach using the true skill statistics (TSS), which is one of the most widely used.

The main idea of the TSS approach is to pick a threshold value and set the cells of the map above to `0` if their values if less than this threshold and to `1` otherwise.
These `0/1` values are then compared to the known occurrence of the species to calculate

  - The number of occurrences correctly predicted as presences (true positives, TP).
  - The number of occurrences incorrectly predicted as absences (false negatives, FN).
  - The number of occurrences correctly as absences (true negatives, TN).
  - The number of occurrences incorrectly presences (false positives, FP).

TSS is defined as $TSS = \frac{TP\,TN - FP\,FN}{(TP+FN)(TN+FP)}$, which is a statistic balancing how well the model performs in predicting both presences and absences.
TSS ranges from 0, for a model not better than random, to 1, for a model with perfect predictions.

If we pick several threshold and calculate the TSS for each of them, the best threshold is the one that has highest TSS, which is also the TSS of our model predictions.

```{r sdm-binary-dummy, fig.width=5, fig.height=4, fig.align='center', eval=FALSE}
# extract the values from the continuous map
suit <- extract(sdm, d[, c("x", "y")], ID = FALSE)[, 1]

# generate a gradient of threshold values
threshold <- seq(0.1, 0.9, by = 0.001)
tss <- rep(NA, length(threshold)) # empty vector for storage

# iterate over threshold values
for (i in seq_along(threshold)) {
  p <- ifelse(suit > threshold[i], 1, 0)
  TP <- sum(p == 1 & d$occ == 1)
  FP <- sum(p == 1 & d$occ == 0)
  FN <- sum(p == 0 & d$occ == 1)
  TN <- sum(p == 0 & d$occ == 0)
  sens <- TP / (TP + FN)
  spec <- TN / (TN + FP)
  tss[i] <- sens + spec - 1
}

th <- threshold[which.max(tss)] # best threshold

plot(
  threshold, tss,
  type = "l", main = paste0("Highest TSS = ", round(max(tss), 2)),
  xlab = "Threshold value", ylab = "TSS"
)
abline(v = th, lty = 2)

# binarize the continuous map
sdm_bin <- ifel(sdm >= th, 1, 0)
plot(sdm_bin, col = c("grey90", "dodgerblue"))
```

```{r sdm-binary, fig.width=5, fig.height=4, fig.align='center', echo=FALSE}
ff <- list.files("../data", pattern = ".tif") # all files with .tif extension
r <- rast(file.path("..", "data", ff))
roi <- ext(-13, 33, 33, 62) # roi of Europe
r <- crop(r, roi) # crop to Europe
sdm <- predict(r, full_enm, type = "response")

# extract the values from the continuous map
suit <- extract(sdm, d[, c("x", "y")], ID = FALSE)[, 1]

# generate a gradient of threshold values
threshold <- seq(0.1, 0.9, by = 0.001)
tss <- rep(NA, length(threshold)) # empty vector for storage

# iterate over threshold values
for (i in seq_along(threshold)) {
  p <- ifelse(suit > threshold[i], 1, 0)
  TP <- sum(p == 1 & d$occ == 1)
  FP <- sum(p == 1 & d$occ == 0)
  FN <- sum(p == 0 & d$occ == 1)
  TN <- sum(p == 0 & d$occ == 0)
  sens <- TP / (TP + FN)
  spec <- TN / (TN + FP)
  tss[i] <- sens + spec - 1
}

th <- threshold[which.max(tss)] # best threshold

plot(
  threshold, tss,
  type = "l", main = paste0("Highest TSS = ", round(max(tss), 2)),
  xlab = "Threshold value", ylab = "TSS"
)
abline(v = th, lty = 2)

# binarize the continuous map
sdm_bin <- ifel(sdm >= th, 1, 0)
plot(sdm_bin, col = c("grey90", "dodgerblue"))
```

This binary map is quite incorrect for this species.
According to the IUCN (<https://www.iucnredlist.org/species/61550/12514105>), this species is found also in most of Italy, all the Balkans, and part of Turkey, but is not found in the UK, the Low Countries, and most of Germany.
Why do we get such bad projections compared to the known range from IUCN?
We will answer this in a next lecture, but my general recommendation is to plot the detection points.

```{r sdm-issue-dummy, fig.width=5, fig.height=4, fig.align='center', eval=FALSE}
plot(sdm_bin, col = c("grey90", "dodgerblue"))
points(d[d$occ == 1, c("x", "y")], pch = 20, cex = .3)
```

```{r sdm-issue, fig.width=5, fig.height=4, fig.align='center', echo=FALSE}
ff <- list.files("../data", pattern = ".tif") # all files with .tif extension
r <- rast(file.path("..", "data", ff))
roi <- ext(-13, 33, 33, 62) # roi of Europe
r <- crop(r, roi) # crop to Europe
sdm <- predict(r, full_enm, type = "response")
th <- threshold[which.max(tss)] # best threshold
sdm_bin <- ifel(sdm >= th, 1, 0)

plot(sdm_bin, col = c("grey90", "dodgerblue"))
points(d[d$occ == 1, c("x", "y")], pch = 20, cex = .3)
```

The selected detection records are geographically biased and likely the environmental (climatic) space is not well represented.
