---
title: "Fitting ENMs"
subtitle: "Generalized Additive Models (GAMs)"
author: ""
format:
  beamer:
    engine: tectonic
    colortheme: beaver
    incremental: false
    aspectratio: 169
---

## Duality of SDMs
\begin{minipage}[t]{0.48\textwidth}
\centering
\textbf{Ecological niche model (ENM)}

\includegraphics[height=5cm]{../figures/niche-1d.pdf}

Ecological space
\end{minipage}
\pause
\begin{minipage}[t]{0.48\textwidth}
\centering
\textbf{Species distribution model (SDM)}

\includegraphics[height=5cm]{../figures/niche-1d-sdm.png}

Geographic space
\end{minipage}

## Ecological Niche Model (ENM)
\begin{minipage}[t]{0.48\textwidth}
\centering
\textbf{Ecological niche model (ENM)}

\includegraphics[height=5cm]{../figures/niche-1d.pdf}

Ecological space
\end{minipage}
\begin{minipage}[t]{0.48\textwidth}
\centering
\end{minipage}

# Introduction to GAMs

## Generalized additive models (GAMs)

 - Machine learning statistical framework
 - Highly flexible and extremely powerful
 - Easy to interpret
 - Easy to implement in R (`mgcv` \& `gratia`)

## Example -- Fitting a 1D niche

```{r 1D, fig.height=2, fig.width=3, fig.align='center'}
library(mgcv)
library(gratia)
d <- read.csv("../data/niche-1d.csv")
enm <- gam(suitability ~ s(temperature, k = 20), data = d)
draw(enm, residuals = TRUE, rug = FALSE)
```

## Example -- Fitting a 2D niche

\begin{center}
\includegraphics[width=6cm]{../figures/niche-2d.png}
\end{center}

## Example -- Fitting a 2D niche

```{r 2D, fig.height=1.8, fig.width=6, fig.align='center'}
d <- read.csv("../data/niche-2d.csv")
enm <- gam(
  suitability ~ s(temperature, k = 20) + s(precipitation, k = 20) +
                ti(temperature, precipitation, k = 16),
  data = d
)
draw(enm, residuals = TRUE, rug = FALSE, nc = 3)
```

## Example -- Fitting a 2D niche

::::{.columns}

:::{.column width=50%}
```{r, fig.height=2, fig.width=3, fig.align='center', echo=FALSE}
draw(enm, residuals = TRUE, rug = FALSE, select = 3)
```
:::

\pause

:::{.column width=50%}
```{r raster, fig.height=3, fig.width=3, fig.align='center', echo=FALSE}
niche <- predict(enm, type = "response")
image(
  d$temperature[1:50],
  d$precipitation[seq(1, nrow(d), by = 50)],
  matrix(d$suitability, 50, 50),
  xlab = "Temperature (C)",
  ylab = "Precipitation (mm)",
  col = hcl.colors(100, "Spectral", rev = TRUE),
  useRaster = TRUE,
  las = 1
)
contour(
  d$temperature[1:50],
  d$precipitation[seq(1, nrow(d), by = 50)],
  matrix(niche, 50, 50),
  add = TRUE,
  levels = seq(0.1, 1, by = 0.2),
  col = "grey20",
  lwd = 2
)
```
:::

::::

## GAM workflow

 1. Fit GAM model: `gam(<y> ~ <x>)`
 \pause
 2. Simple covariate: `s(<x>, k = ...)`
 \pause
 3. Interaction term: `ti(x1, x2, k = ...)`
 \pause
 4. Inspect: `draw(...)`
 \pause
 5. Appraise: `k.check(...)`
    - $k-index \ll 1 \rightarrow \text{Increase} k$
    - $p-value < 0.05 \rightarrow \text{Increase} k$

# Data for ENM/SDM

## GAM -- SDM data

 - Rarely (if ever) you will get suitability as data
 - More likely (almost always) you will get presence (1) and absence (0) data

::::{.columns}

:::{.column width=50%}
```{r, fig.height=3, fig.width=3, fig.align='center', echo=FALSE}
subd <- d[1:50 + 450, ]
plot(
  subd$suitability, subd$occurrence,
  bg = "dodgerblue3", pch = 21,
  xlab = "Suitability", ylab = "Occurrence"
)
lines(
  sort(subd$suitability), sort(1 / (1 + 100 * exp(-10 * subd$suitability))),
  col = "grey20", lwd = 3
)
```
:::

::::

## GAM -- SDM data

 - Rarely (if ever) you will get suitability as data
 - More likely (almost always) you will get presence (1) and absence (0) data

::::{.columns}

:::{.column width=50%}
```{r, fig.height=3, fig.width=3, fig.align='center', echo=FALSE}
subd <- d[1:50 + 450, ]
plot(
  subd$suitability, subd$occurrence,
  bg = "dodgerblue3", pch = 21,
  xlab = "Suitability", ylab = "Occurrence"
)
lines(
  sort(subd$suitability), sort(1 / (1 + 100 * exp(-10 * subd$suitability))),
  col = "grey20", lwd = 3
)
```
:::

:::{.column width=50%}
```{r, fig.height=3, fig.width=3, fig.align='center', echo=FALSE}
plot(
  subd$temperature, subd$suitability,
  type = "l", lwd = 3, col = "dodgerblue3",
  xlab = "Temperature (C)", ylab = "Suitability",
  xaxs = "i", yaxs = "i", ylim = c(-0.01, 1.1)
)
```
:::

::::

## GAM -- SDM data

 - Rarely (if ever) you will get suitability as data
 - More likely (almost always) you will get presence (1) and absence (0) data

::::{.columns}

:::{.column width=50%}
```{r, fig.height=3, fig.width=3, fig.align='center', echo=FALSE}
subd <- d[1:50 + 450, ]
plot(
  subd$suitability, subd$occurrence,
  bg = "dodgerblue3", pch = 21,
  xlab = "Suitability", ylab = "Occurrence"
)
lines(
  sort(subd$suitability), sort(1 / (1 + 100 * exp(-10 * subd$suitability))),
  col = "grey20", lwd = 3
)
```
:::

:::{.column width=50%}
```{r, fig.height=3, fig.width=3, fig.align='center', echo=FALSE}
plot(
  subd$temperature, subd$suitability,
  type = "l", lwd = 3, col = "dodgerblue3",
  xlab = "Temperature (C)", ylab = "Suitability",
  xaxs = "i", yaxs = "i", ylim = c(-0.01, 1.1)
)
lines(
  subd$temperature, (1 / (1 + 100 * exp(-10 * subd$suitability))),
  col = "grey20", lwd = 3
)
```
:::

::::

## GAM -- SDM data

 - Rarely (if ever) you will get suitability as data
 - More likely (almost always) you will get presence (1) and absence (0) data

::::{.columns}

:::{.column width=50%}
```{r, fig.height=3, fig.width=3, fig.align='center', echo=FALSE}
subd <- d[1:50 + 450, ]
plot(
  subd$suitability, subd$occurrence,
  bg = "dodgerblue3", pch = 21,
  xlab = "Suitability", ylab = "Occurrence"
)
lines(
  sort(subd$suitability), sort(1 / (1 + 100 * exp(-10 * subd$suitability))),
  col = "grey20", lwd = 3
)
```
:::

:::{.column width=50%}
```{r, fig.height=3, fig.width=3, fig.align='center', echo=FALSE}
plot(
  subd$temperature, subd$suitability,
  type = "l", lwd = 3, col = "dodgerblue3",
  xlab = "Temperature (C)", ylab = "Suitability",
  xaxs = "i", yaxs = "i", ylim = c(-0.01, 1.1)
)
lines(
  subd$temperature, (1 / (1 + 100 * exp(-10 * subd$suitability))),
  col = "grey20", lwd = 3
)
occ <- rbinom(nrow(subd), 1, (1 / (1 + 100 * exp(-10 * subd$suitability))))
points(
  subd$temperature, occ, 
  pch = 21, bg = ifelse(occ == 0, "white", "goldenrod")  
)
```
:::

::::

## GAM -- Example with SDM data

```{r, fig.height=1, fig.width=2, fig.align='center'}
d <- read.csv("../data/niche-1d.csv")
enm <- gam(occurrence ~ s(temperature, k = 30), data = d, family = binomial)
# draw(enm, rug = FALSE, residuals = TRUE)
library(ggplot2)
fitted_values(enm) |>
  ggplot() +
  geom_point(data = d, aes(temperature, suitability), alpha = .3, col = "dodgerblue3") +
  geom_ribbon(aes(x = temperature, ymin = .lower_ci, ymax = .upper_ci), alpha = 0.2) +
  geom_line(aes(temperature, .fitted))
```

## GAM -- Fitted niche is scaled

```{r, fig.height=2, fig.width=2, fig.align='center'}
fitted_values(enm) |>
  ggplot() +
  geom_point(data = d, aes(temperature, suitability), alpha = .3, col = "dodgerblue3") +
  geom_ribbon(aes(x = temperature, ymin = .lower_ci / max(.fitted), ymax = .upper_ci / max(.fitted)), alpha = 0.2) +
  geom_line(aes(temperature, .fitted / max(.fitted)))
```

## GAM -- 2D example with SDM data

```{r}
d <- read.csv("../data/niche-2d.csv")
enm <- gam(
  occurrence ~ s(temperature, k = 10) + s(precipitation, k = 10) +
               ti(temperature, precipitation, k = 10),
  data = d,
  family = binomial
)
niche <- predict(enm, type = "response")
```

```{r, fig.height=3, fig.width=3, fig.align='center', echo=FALSE}
image(
  d$temperature[1:50],
  d$precipitation[seq(1, nrow(d), by = 50)],
  matrix(d$suitability, 50, 50),
  xlab = "Temperature (C)",
  ylab = "Precipitation (mm)",
  col = hcl.colors(100, "Spectral", rev = TRUE),
  useRaster = TRUE,
  las = 1
)
contour(
  d$temperature[1:50],
  d$precipitation[seq(1, nrow(d), by = 50)],
  matrix(niche, 50, 50),
  add = TRUE,
  levels = seq(0.1, 1, by = 0.2),
  col = "grey20",
  lwd = 2
)
```

# Model selection

## AIC

You can perform model selection on ENMs just as any other statistical models.

```{r}
enm_simple <- gam(
  occurrence ~ s(temperature, k = 10) + s(precipitation, k = 10),
  data = d, family = binomial, method = "ML"
)
enm_int <- gam(
  occurrence ~ s(temperature, k = 10) + s(precipitation, k = 10) +
               ti(temperature, precipitation, k = 10),
  data = d, family = binomial, method = "ML"
)
AIC(enm_simple, enm_int)
```

## Maximum Likelihood

When you perform model selection using AIC, remember to specify `gam(..., method = "ML")`.

Once you select the model, you can update it and remove `method = "ML"`.
